---
title: "A13-rPython"
author: "Verena Haunschmid"
date: "14 September 2016"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## rPython

Recently I have started using `Python` for data analysis. By coincidence I saw that someone on Twitter mentioned the R package [rPython](https://cran.r-project.org/web/packages/rPython/rPython.pdf). Since sometimes it's convenient to mix languages in a project, I wanted to know more about it.

[Here](https://rpython.readthedocs.io/en/latest/) you can find lots of documentation.

### Install & load rPython

```{r, eval=FALSE}
install.packages("rPython")
```

```{r}
library(rPython)
```
### The package & methods

The package contains four main methods:

- `python.assign`
- `python.call`
- `python.exec`
- `python.load`

It also contains another method:

- `python.get`

If you call any of those methods it saves it to some internal workspace (I don't know any details about that but you can probably find out in the link above).

## Using the package

In the following I'll show examples for all 4 methods.

```{r, echo=FALSE}
python.exec('def cool_python_fct(a, b, c): print "magic python stuff"; return a + b + c')
```

### python.assign

If we want to assign some values in our magic *environment* we can use this method.

```{r}
python.assign('a', 5)
python.assign('b', 2)
python.assign('c', 3)
```

For my examples I will also assign the same values to variables with other names. Of course I could use the same names because this is a different workspace but I want the code below to be less confusing.

```{r}
d <- 5
e <- 2
f <- 3
```

### python.call

Assume I have a `python` function named `cool_python_fct` that I want to use but it takes too long to rewrite it in R. It takes 3 variables that I have in my R environment. There is a neat way to use the method anyway:

```{r}
python.call('cool_python_fct', d, e, f)
```

As you can see the result is printed directly!

We can also store it in an R variable.

```{r}
res <- python.call('cool_python_fct', d, e, f)
res
```


### python.exec

There is another way to call/execute a function. If I don't want to pass any R objects, this is also fine. Note that you only get the printed output but if you want to save the computed value you need to assign it to a variable. With `python.get` you can extract it.

```{r}
python.exec('x = cool_python_fct(a, b, c)')
python.get('x')
```

This way I can also do other cool things, like importing python libraries. In the next example we check which python version is used:

```{r}
python.exec("import sys")
python.exec("print sys.version")
```

Interestingly this is only printed when called in the R command line but not when I create the report. But this way it also works:
```{r}
python.get("sys.version")
```

### python.load

This method is really useful to load and execute code in a file. Let's say I have a file that contains all the packages I want to import.

```{r}
python.load("imports.py")
```
Now you can use the `math` package.

```{r}
python.get("math.pi")
```

## What this means

Although the package only has 4 methods, it is actually really mighty. With `python.exec` and `python.call` we have two really powerful tools to combine R and Python. If you prefer to wrangle your data in R but want to use algorithms that are only available in Python, it's super easy now.