---
title: "summarize-data"
author: "Verena Haunschmid"
date: "14 April 2016"
output: pdf_document
---

This document should give a short overview over the different datasets I use in the [Data Science Learning Club](http://becomingadatascientist.com/learningclub/). I'll try to keep it short and don't want to include more than one or two plots per dataset.

Download link, collected by

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

These are libraries I use for visualisation.

```{r libs}
library(tabplot) # tableplot (for categorical data)
library(ggplot2) # general plotting function
library(leaflet)
```

`tabplot` can be installed this way:

```{r, eval=FALSE}
library(devtools)
install_github("mtennekes/tabplot", subdir="pkg")
```

`leaflet` can be installed this way:
```{r}
library(devtools)
install_github("rstudio/leaflet")
```

## Mushroom data

The mushroom dataset is an interesting dataset because it only consists of categorical features (22 to be exact). The target is another categorical value (edible or poisonous).

```{r mush, cache=TRUE}
mush_data<-read.csv("https://archive.ics.uci.edu/ml/machine-learning-databases/mushroom/agaricus-lepiota.data", sep=",", header=FALSE)
```

The `tableplot` ([suggested on Twitter by hrbrmstr](https://twitter.com/hrbrmstr/status/713705491604451328)) is a very useful tool to visualise this dataset.

```{r tableplot}
tableplot(mush_data)
```

## Tractive data

I have written about [this dataset on my blog](http://blog.haunschmid.name/data-collected-with-my-tractive-gps-pet-tracker/) already. It's also available [on my github repo tractive_data](https://github.com/expectopatronum/tractive_data).

```{r}
track_cat_clean <- read.csv("01-summarize-data/data_sets/track_cat_clean.csv")
mycats <- unique(track_cat_clean$cat)
cols <- c("brown", "purple", "blue", "green")
catMap<-leaflet() %>% addTiles()
for (i in 1:length(mycats)) {
  cat<-mycats[i]
  if (cat != "Unknown") {
    catMap<-catMap %>% addPolylines(track_cat_clean[track_cat_clean$cat==cat,"lon"], track_cat_clean[track_cat_clean$cat==cat,"lat"], col=cols[i], group=cat, weight = 2, fillOpacity = 0.4) 
  }
}
 
catMap %>%
addLayersControl( 
  overlayGroups = mycats, 
  options = layersControlOptions(collapsed = FALSE))
```

## Geocaching logs

wordcloud

## Tatort

## NFL data

Other datasets I used for the learning club
